Undecidability and the structure of the Turing degrees

### 2018-08-17 paper.tex
First complete draft!

\section{Introduction}
Historical overview of the field, motivation for study
\cite{ambos-spies06:_degrees_unsol}
\cite{soare16_turin_comput}
\cite{soare1999history}
\cite{lerman16:_degrees_unsol}
\todo{}
\subsection{Decision Problems}
What does it mean to \emph{solve a problem}? What is a \emph{problem}, anyways?
These questions may seem superfluous, they must be addressed in order to build a
well defined theory of the difficulty and computability of problems. In this paper, we discuss a
specific type of problem, the \emph{decision problem}. Decision problems are
essentially sets of natural numbers, and to solve the problem is to be able to
decide whether any given element is inside the set or not. Because set theory is
commonly used as the foundational system of mathematics, we can convert problems
expressed in terms of more complex mathematical objects into decision problems,
and thus by studying decision problems can build a system that is able to assess
the computability of all mathematical problems. For example, we can express
strings over the alphabet $\{0,1\}$ as sets of natural numbers, and thus can
study problems about languages of such strings with the same tools we use for
decision problems. Going further, we can express the source code of a computer
program as a string over the alphabet $\{0,1\}$, and so can consider questions
about the behavior of computer programs.

@@

\begin{prob}[The Halting Problem]
  \label{prob:halting}
  Let $M$ be a Turing Machine and $i$ be an input. If $M$ is run on $i$, will it
  eventually halt?
  \begin{proof}[Proof of Undecidability]
    Suppose for the sake of contradiction that the halting problem is
    decidable. Then, there exists some machine
    $\mathcal{O}$ that can decide the halting problem. Then, we can construct a
    Turing machine $H$ that simulates $\mathcal{O}$ on its input $(M,i)$. In the
    case where $\mathcal{O}(M,i)$ accepts, $H$ enters into an infinite
    loop, whereas in the case where $\mathcal{O}(M,i)$ does not accept, $H$
    halts. Consider $H(H,\epsilon)$. If $H$ halts, then
    $\mathcal{O}(H,\epsilon)$ accepts, and so $H$ enters an infinite loop, and
    so it does not halt. But if $H$ does not halt, then
    $\mathcal{O}(H,\epsilon)$ does not accept, so $H$ halts. This is a
    contradiction, so our premise that the halting problem is decidable must be false.
  \end{proof}
\end{prob}
% \begin{thm}[Rice's Theorem] \cite{kozen99_autom}
%   % Proof by reduction to the halting problem?
% \end{thm}
% \subsection{Incompleteness}
% \begin{thm}[G\"odel's Incompleteness Theorem] 
%   % Kleene 1943? Also shown in hopcroft 1979
%   \cite{kleene43_recur_predic_quant}
% \end{thm}
\subsection{Reducibility}
\label{subsec:reducibility}
While it is possible to prove that a problem is undecidable directly, as in
problem~\ref{prob:halting}, it is often more convenient to prove undecidability
through comparison to problems which are already known to be undecidable. This
comparison takes place through the technique of Turing reduction.
\begin{defn}
  Let $A$ and $B$ be decision problems. We say that $A$ is \emph{Turing reducible}
  to $B$ and write $A \leq_T B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
  a machine $D_A$ that decides $A$.
\end{defn}
So, we have that if $A$ is reducible to $B$, then $A$ can be no harder than $B$,
because any solution to $B$ also leads to a solution of $B$. So, if we have some
problem $B$ that we would like to prove is undecidable, we can do so by showing
that some problem $A$ which is already known to be undecidable is reducible to
$B$. Since $A$ cannot be harder than $B$, it follows that $B$ must also be
undecidable. Alternatively, if we would like to show that some problem $A$ is
decidable, it is sufficient to show that it is reducible to some problem $B$
that is known to be decidable.
\cite{sipser13:_introd_theor_comput}
\cite{post44:_recur}

\cite{kleene80_introd}

@@

\begin{thm}[Closure Properties of \acp{CFL}]
  \aclp{CFG} have a variety of closure properties:
  \begin{itemize}
  \item Let $L_1,L_2$ be \acp{CFL}. Then $L_1\cup L_2$ is a \ac{CFL}.
    \begin{proof}
      The languages $L_1$ and $L_2$ are \aclp{CFL}, so there exist \aclp{PDA} $P_1$ and $P_2$
      that decide them. Consider the \ac{PDA} $U$ shown in
      Figure~\ref{pda:union}, which simulates $P_1$ and $P_2$ on the input. $U$
      is a \ac{PDA} that decides $L_1\cup L_2$, so it follows that
      $L_1\cup L_2$ is a \ac{CFL}.
      \begin{figure}[h]
        \caption{$U$, a \ac{PDA} that simulates $P_1$ and $P_2$.}
        \label{pda:union}
         \begin{tikzpicture}[shorten >=1pt,node distance=3cm,on grid,auto]
  \node[state,initial] (q_0) {$q_0$};
  \node[state,diamond] (P_1) [above right=of q_0]{$P_1$};
  \node[state,diamond] (P_2) [below right=of q_0]{$P_2$};
  \draw[->]
  (q_0) edge node {$\epsilon,\epsilon\to\epsilon$} (P_1);
  \draw[->]
  (q_0)
  edge node {$\epsilon,\epsilon\to\epsilon$} (P_2);
  \end{tikzpicture}
      \end{figure}
    \end{proof}
  \item Let $L_1,L_2$ be \acp{CFL}. Then the concatenation \[
      \{w_1w_2 \mid w_1\in L_1, w_2 \in L_2\}
    \]
    is a \ac{CFL}.
    \begin{proof}
      The languages $L_1$ and $L_2$ are \aclp{CFL}, so there exist \acp{CFG}
      $G_1,G_2$ with start symbols $S_1,S_2$ that produce them. Suppose, without loss of generality, that
      $G_1,G_2$ have no nonterminal symbols in common (if they do, then we can rename
      them). Consider the grammar $S \rightarrow S_1S_2$. Then, this is a
      context free grammar, and its language is the concatenation of $L_1$ and
      $L_2$. So, the concatenation of $L_1$ and $L_2$ is a \ac{CFL}.
    \end{proof}
  \item Let $L$ be a \ac{CFL}. Then,
    \[
      L^* = \bigcup_{i \geq 0}\{w_1 \dots w_k\dots w_i \mid w_k\in L\}
    \]
    is a \ac{CFL}.
    \begin{proof}
      The language $L$ is a \ac{CFL}, so there exists a \ac{CFG} $G$ with start
      symbol $S_1$ that produces it. Consider the grammar given by $S\rightarrow
      S_1S\mid\epsilon$. This is a \ac{CFG} that produces $L^*$, so it follows
      that $L^*$ is a \ac{CFL}.
    \end{proof}
    \item Intersection with regular languages~\cite{hopcroft07:_introd_autom_theor_languag_comput}.\todo{}
  \end{itemize}
\end{thm}

\begin{prob}[Emptiness]
  \label{prob:cfg:emptiness}
  Let $G$ be a context-free grammar. Is $L(G)=\varnothing$?
  \begin{proof}[Proof of Decidability]
    \todo{}
  \end{proof}
\end{prob}

\begin{prob}[Finite]
  Let $G$ be a context-free grammar. Is $L(G)$ finite?
  \begin{proof}[Proof of Decidability]
    \todo{}
  \end{proof}
\end{prob}

\begin{prob}[Regular Containment]
Take some context-free grammar $G$ and some regular language $R$. Is
$L(G)\subseteq R$?
\begin{proof}[Proof of Decidability]
  We have that $L(G)\subseteq R$ if and only if
  $\overline{L(G)}\cup R = \Sigma^*$. So, it follows that $L(G)$ is contained by
  $R$ if and only iff $\overline{\overline{L(G)}\cup R} = \varnothing$. By
  DeMorgan's laws, we have that this is equivalent to the statement
  $L(G)\cap \overline{R} = \varnothing$. Regular languages are closed under
  complement and context-free languages are closed under intersection with a
  regular language, so it follows that $L(G)\cap \overline{R}$ is a context free
  language. So the problem of containment by a regular language is reducible to
  Problem~\ref{prob:cfg:emptiness}, the problem of emptiness, which is
  decidable. So, the problem of containment by a regular language is decidable.
\end{proof}

@@

\subsection{Undecidable Problems}
\begin{prob}[The Post Correspondence Problem]
  Consider some alphabet $\Sigma$ and two finite lists of words over $\Sigma$
  denoted $A = a_1,\dots,a_n$ and $B = b_1, \dots, b_n$. Then, does there exist
some sequence of indices $(i_k),$ for $1 \leq k \leq K$ and for
  $K\geq 1$, and with $1 \leq i_k \leq n$ for all $k$ such that
  \[
    a_{i_1}\dots a_{i_K} = b_{i_1}\dots b_{i_K}?
  \]
  \begin{proof}[Solution]
    This problem is shown to be undecidable through a reduction to the halting
    problem. The proof is technical, and involves encoding the computation
    history of a Turing machine on an input in such a way that the encoding satisfies the
    PCP if and only if the Turing machine accepts the input. It is described in
    detail by Sipser in his book \emph{Introduction to the Theory of Computation}
    \cite{sipser13:_introd_theor_comput}.
  \end{proof}
\end{prob}

@@

\begin{prob}[Disjointness]
  Let $P,Q$ be context-free grammars. Is $L(P)\cap L(Q) = \varnothing$?
  \begin{proof}[Proof of Undecidability]
    Consider the Post correspondence problem for two lists of words $A,B$. We
    construct two context free grammars from these lists as follows:
    \begin{equation*}
      \begin{split}
        G_A \rightarrow  &\;a_11 \\
        &\!\!\!\!\vdots \\
        G_A \rightarrow  &\;a_nn \\
        G_A \rightarrow  &\;a_1G_A1 \\
        &\!\!\!\!\vdots \\
        G_A \rightarrow  &\;a_nG_An \\
      \end{split}
      \qquad
      \begin{split}
        G_B \rightarrow  &\;a_11 \\
        &\!\!\!\!\vdots \\
        G_B \rightarrow  &\;a_nn \\
        G_B \rightarrow  &\;a_1G_B1 \\
        &\!\!\!\!\vdots \\
        G_B \rightarrow  &\;a_nG_Bn \\
      \end{split}
    \end{equation*}
    Then, we can observe that for some string $s$ to exist in both $L(G_A)$ and $L(G_B)$,
    it must be a solution to the Post correspondence for $A,B$. So, if $L(G_A)\cap
    L(G_B)=\varnothing$, then there are no solutions to the Post correspondence
    for $A,B$. So, it follows that the Post correspondence problem is reducible
    to the problem of the disjointness of context-free grammars, so since the
    Post correspondence problem is undecidable, it follows that the disjointness
    problem is undecidable.
  \end{proof}
\end{prob}

\begin{prob}[Universality]
  \label{prob:cfg:universality}
  Let $G$ be some context-free grammar over an alphabet $\Sigma$. Then, is
  \[
    L(G) = \Sigma^*?
  \]
  \begin{proof}[Proof of Undecidability]\todo{}

  \end{proof}
\end{prob}

@@

\begin{prob}[Ambiguity]
  Let $G$ be a context-free grammar. Is $G$ ambiguous?
  \begin{proof}[Proof of Undecidability]
    Consider the Post correspondence problem for two lists of words
    $A,B$. Construct their corresponding grammars $G_A,G_B$. Now, consider the
    grammar
    \[
      G \rightarrow G_A \vert G_B.
    \]
    It follows that the ambiguity of $G$ implies that a solution to the Post
    correspondence problem for $A,B$ exists, so the Post correspondence problem
    reduces to the ambiguity problem, so the ambiguity problem is undecidable.
  \end{proof}
\end{prob}
\cite{greibach66:_unsol_recog_linear_contex_free_languag}
\cite{Hopcroft1969}
\begin{prob}
  Let $G$ be a context free grammars. Is $L(G)$ regular?
  \begin{proof}[Proof of Undecidability]
    Note that $\Sigma^*$ is regular, so if $L(G)$ is not regular then it follows
    that $L(G)\neq \Sigma^*$. Additionally, universality is decidable within
    regular languages, because they are closed under complement and emptiness is
    decidable even within context-free grammars. So, we have that
    Problem~\ref{prob:cfg:universality} reduces to the regularity problem. It
    follows that the regularity problem is undecidable.
  \end{proof}
\end{prob}
\begin{prob}[Equality]
  Let $G_1,G_2$ be context free grammars. Is $L(G_1)=L(G_2)$?
  \begin{proof}[Proof of Undecidability]
    Note that $\Sigma^*$ is regular, so it is also a context-free. So,
    we have that Problem~\ref{prob:cfg:universality} reduces to the equality
    problem. It follows that the equality problem is undecidable.
  \end{proof}
\end{prob}
\begin{prob}[Inclusion]
  Let $G_1,G_2$ be context free grammars. Is $L(G_1)\subseteq L(G_2)$?
  \begin{proof}[Proof of Undecidability]
    As before, note that $\Sigma^*$ is regular, so it is also a
    context-free. Additionally, observe that for any grammar $G$, $L(G)\subseteq
    \Sigma^*$. So,
    we have that Problem~\ref{prob:cfg:universality} reduces to the inclusion
    problem, since $\Sigma^* \subseteq L(G) \implies \Sigma^* = L(G)$. It follows that the inclusion problem is undecidable.
  \end{proof}
\end{prob}

\subsubsection{Problems decidable for DCFLs}
\cite{ginsburg65:_deter}
\begin{defn}
  A \ac{DCFL} is a language accepted by a
  \acl{DPDA}.
\end{defn}

@@

\begin{prob}
  Let $L$ be a \ac{DCFL}. Is $L = \Sigma^*$?
  \begin{proof}[Proof of Decidability]
    We have that $L$ is a \ac{DCFL}, so by Theorem~\ref{dcfl:complement},
    $\overline{L}$ is also a $\ac{DCFL}$. By Problem~\ref{prob:cfg:emptiness},
    we have that whether $\overline{L}$ is empty is decidable, so it follows
    that the question of whether $L = \Sigma^*$ is also decidable.
  \end{proof}
\end{prob}

@@

One of the earliest questions about the structure of the Turing degrees was
Post's Problem, first posed in 1944~\cite{post44:_recur}. It was solved
independently by Friedberg and Mucnik in the late 1950s using a technique
called the ``priority method''~\cite{Friedberg236, muchnik1956unsolvability}. A solution that does not
involve a priority argument was provided by Ku\v{c}era in 1986~\cite{Kucera:1986:APS:22416.22462}.

\begin{prob}[Post's Problem]  
  Does there exist a computably enumerable degree strictly between $0$ and $0'$?
  % Marchenkov 1976
  % Independently by Friedberg 57, Mucnik 56/58 using priority method
  % Priority free solution Kucer 1986''
  % ``Natural'' solns don't exist
\end{prob}

@@

\begin{lem}
  \label{lem:deg_count}
  Every Turing degree is of cardinality $\aleph_0$.
  \begin{proof}
    We have that since every Turing machine can be expressed as a finite string over
    $\{0,1\}$, it can be expressed as a finite subset of $\mathbb{N}$. So, it follows
    that for any given oracle, there are at most $\aleph_0$ oracle machines, so
    at most $\aleph_0$ problems are decidable for any Turing degree.
    \par
    Now that an upper bound has been established,  we show that the cardinality
    of each degree is exactly $\aleph_0$.\todo{}
  \end{proof}
\end{lem}

@@

\section{Further Reading}
The study of the Turing degrees was a very active topic in the late
20\textsuperscript{th} and early 21\textsuperscript{st} centuries, and this
paper has just scratched the surface of the field. \par
Readers particularly interested in the structure of the computably enumerable
degrees might be interested in a result by Sacks in 1964 showing their
density~\cite{sacks64:_recur_enumer_degrees_dense}, while those interested in
\aclp{CFG} or the Turing degrees of problems about them should consult Reedy's
1975 paper on the Turing degree of a specific problem~\cite{REEDY197577} and
Ginsburg's book \emph{The Mathematical Theory of Context Free
  Languages}~\cite{ginsburg66:_mathem_theor_contex_languag}. Additionally, a
simple overview of the field is available in Hopcroft's classic text on
computability
theory~\cite[chs.~5--7]{hopcroft07:_introd_autom_theor_languag_comput}. \par For excellent, comprehensive,
and more rigorous textbooks on the study of Turing computability and the Turing
degrees, consider Soare and Lerman's
books~\cite{soare16_turin_comput,lerman16:_degrees_unsol}, as well as
Kleene's original text on the subject~\cite{kleene80_introd}.

>>>>

\section{Introduction}

The field of computability theory focuses on abstract models of computation and
the sorts of problems that they are able to solve. Unlike complexity theory,
which is focused on the resources required to perform a computation,
computability theory focuses only on whether a certain computation is possible
on a given machine. The practical application of the theory focuses on models of
computation of equivalent power to general purpose computers, allowing for it to
be determined whether it possible to solve specific problems on actual
machines. The Church-Turing thesis, a significant conjecture in the field,
states that the model that we use to understand the power of general purpose
computers has the same power as humans applying algorithms. If true, it means
that it is not possible to design a computational process that is able to solve
more problems than a general-purpose computer. However, even if it is true, it is still possible to
imagine models that have more power than general purpose computers by endowing
them with \emph{oracles} the ability to evaluate problems that we know
general-purpose computers cannot. Because of this, computability is still able
to draw distinctions in the difficulty of problems too difficult for
general-purpose computers to solve, dividing them into \emph{Turing degrees}, or
\emph{degrees of unsolvability}. Additionally, it is possible to construct
models of computation with less power than general-purpose computers, and thus
classify the difficulty of problems that general-purpose computers can solve. We
can also show the differences in difficulty of problems simply by showing that
it is possible to use the answer of one problem to solve another, a technique
called \emph{reduction}. So, it is possible to create an infinite hierarchy
describing the difficulty of problems, and to investigate the sort of structure
this hierarchy takes.

% Historical overview of the field, motivation for study

% \cite{soare16_turin_comput}
% \cite{soare1999history}
% \cite{lerman16:_degrees_unsol}
% \todo{}
\subsection{Decision Problems}
What does it mean to \emph{solve a problem}? What is a \emph{problem}, anyways?
These questions may seem esoteric, they must be addressed in order to build a
well defined theory of the difficulty and computability of problems. In this paper, we discuss a
specific type of problem, the \emph{decision problem}. Decision problems are
essentially sets of natural numbers, and to solve the problem is to be able to
decide whether any given element is inside the set or not. Because set theory is
commonly used as the foundational system of mathematics, we can convert problems
expressed in terms of more complex mathematical objects into decision problems,
and thus by studying decision problems can build a system that is able to assess
the computability of all mathematical problems. For example, we can express
strings over the alphabet $\{0,1\}$ as sets of natural numbers, and thus can
study problems about languages of such strings with the same tools we use for
decision problems. Going further, we can express the source code of a computer
program as a string over the alphabet $\{0,1\}$, and so can consider questions
about the behavior of computer programs.

@@

\begin{prob}[The Halting Problem]
  \label{prob:halting}
  Let $M$ be a Turing Machine and $i$ be an input. If $M$ is run on $i$, will it
  eventually halt?
  \begin{proof}[Proof of Undecidability]
    Suppose for the sake of contradiction that the halting problem is
    decidable. Then, there exists some machine
    $\mathcal{O}$ that can decide the halting problem. Then, we can construct a
    Turing machine $H$ that simulates $\mathcal{O}$ on its input $(M,i)$. In the
    case where $\mathcal{O}(M,i)$ accepts, $H$ enters into an infinite
    loop, whereas in the case where $\mathcal{O}(M,i)$ does not accept, $H$
    halts. Consider $H(H,\epsilon)$. If $H$ halts, then
    $\mathcal{O}(H,\epsilon)$ accepts, and so $H$ enters an infinite loop, and
    so it does not halt. But if $H$ does not halt, then
    $\mathcal{O}(H,\epsilon)$ does not accept, so $H$ halts. This is a
    contradiction, so our premise that the halting problem is decidable must be false.
  \end{proof}
\end{prob}
% \begin{thm}[Rice's Theorem] \cite{kozen99_autom}
%   % Proof by reduction to the halting problem?
% \end{thm}
% \subsection{Incompleteness}
% \begin{thm}[G\"odel's Incompleteness Theorem] 
%   % Kleene 1943? Also shown in hopcroft 1979
%   \cite{kleene43_recur_predic_quant}
% \end{thm}
\subsection{Reducibility}
\label{subsec:reducibility}
While it is possible to prove that a problem is undecidable directly, as in
problem~\ref{prob:halting}, it is often more convenient to prove undecidability
through comparison to problems which are already known to be undecidable. This
comparison takes place through the technique of Turing reduction.
\begin{defn}
  Let $A$ and $B$ be decision problems. We say that $A$ is \emph{Turing reducible}
  to $B$ and write $A \leq_T B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
  a machine $D_A$ that decides $A$.
\end{defn}
So, we have that if $A$ is reducible to $B$, then $A$ can be no harder than $B$,
because any solution to $B$ also leads to a solution of $B$. So, if we have some
problem $B$ that we would like to prove is undecidable, we can do so by showing
that some problem $A$ which is already known to be undecidable is reducible to
$B$. Since $A$ cannot be harder than $B$, it follows that $B$ must also be
undecidable. Alternatively, if we would like to show that some problem $A$ is
decidable, it is sufficient to show that it is reducible to some problem $B$
that is known to be decidable~\cite{sipser13:_introd_theor_comput,post44:_recur,kleene80_introd}.

@@

\begin{thm}[Closure Properties of \acp{CFL}]
  \aclp{CFG} have a variety of closure properties:
  \begin{itemize}
  \item Let $L_1,L_2$ be \acp{CFL}. Then $L_1\cup L_2$ is a \ac{CFL}.
    \begin{proof}
      The languages $L_1$ and $L_2$ are \aclp{CFL}, so there exist \aclp{PDA} $P_1$ and $P_2$
      that decide them. Consider the \ac{PDA} $U$ shown in
      Figure~\ref{pda:union}, which simulates $P_1$ and $P_2$ on the input. $U$
      is a \ac{PDA} that decides $L_1\cup L_2$, so it follows that
      $L_1\cup L_2$ is a \ac{CFL}.
      \begin{figure}[h]
        \caption{$U$, a \ac{PDA} that simulates $P_1$ and $P_2$.}
        \label{pda:union}
         \begin{tikzpicture}[shorten >=1pt,node distance=3cm,on grid,auto]
  \node[state,initial] (q_0) {$q_0$};
  \node[state,diamond] (P_1) [above right=of q_0]{$P_1$};
  \node[state,diamond] (P_2) [below right=of q_0]{$P_2$};
  \draw[->]
  (q_0) edge node {$\epsilon,\epsilon\to\epsilon$} (P_1);
  \draw[->]
  (q_0)
  edge node {$\epsilon,\epsilon\to\epsilon$} (P_2);
  \end{tikzpicture}
      \end{figure}
    \end{proof}
  \item Let $L_1,L_2$ be \acp{CFL}. Then the concatenation \[
      \{w_1w_2 \mid w_1\in L_1, w_2 \in L_2\}
    \]
    is a \ac{CFL}.
    \begin{proof}
      The languages $L_1$ and $L_2$ are \aclp{CFL}, so there exist \acp{CFG}
      $G_1,G_2$ with start symbols $S_1,S_2$ that produce them. Suppose, without loss of generality, that
      $G_1,G_2$ have no nonterminal symbols in common (if they do, then we can rename
      them). Consider the grammar $S \rightarrow S_1S_2$. Then, this is a
      context free grammar, and its language is the concatenation of $L_1$ and
      $L_2$. So, the concatenation of $L_1$ and $L_2$ is a \ac{CFL}.
    \end{proof}
  \item Let $L$ be a \ac{CFL}. Then,
    \[
      L^* = \bigcup_{i \geq 0}\{w_1 \dots w_k\dots w_i \mid w_k\in L\}
    \]
    is a \ac{CFL}.
    \begin{proof}
      The language $L$ is a \ac{CFL}, so there exists a \ac{CFG} $G$ with start
      symbol $S_1$ that produces it. Consider the grammar given by $S\rightarrow
      S_1S\mid\epsilon$. This is a \ac{CFG} that produces $L^*$, so it follows
      that $L^*$ is a \ac{CFL}.
    \end{proof}
    \item Intersection with regular languages~\cite{hopcroft07:_introd_autom_theor_languag_comput}.
  \end{itemize}
\end{thm}

We have that the emptiness and finiteness problems are decidable for
\acp{CFG}~\cite{hopcroft07:_introd_autom_theor_languag_comput,
  sipser13:_introd_theor_comput}. Another problem that is decidable for
\acp{CFG} is the question of whether a given \ac{CFL} is wholly contained by a
given regular language.

\begin{prob}[Regular Containment]
Take some context-free grammar $G$ and some regular language $R$. Is
$L(G)\subseteq R$?
\begin{proof}[Proof of Decidability]
  We have that $L(G)\subseteq R$ if and only if
  $\overline{L(G)}\cup R = \Sigma^*$. So, it follows that $L(G)$ is contained by
  $R$ if and only iff $\overline{\overline{L(G)}\cup R} = \varnothing$. By
  DeMorgan's laws, we have that this is equivalent to the statement
  $L(G)\cap \overline{R} = \varnothing$. Regular languages are closed under
  complement and context-free languages are closed under intersection with a
  regular language, so it follows that $L(G)\cap \overline{R}$ is a context free
  language. So the problem of containment by a regular language is reducible to
  the problem of emptiness, which we know to be
  decidable. So, the problem of containment by a regular language is decidable.
\end{proof}

@@

\subsection{Undecidable Problems}
\begin{prob}[The Post Correspondence Problem]
  \label{prob:pcp}
  Consider some alphabet $\Sigma$ and two finite lists of words over $\Sigma$
  denoted $A = a_1,\dots,a_n$ and $B = b_1, \dots, b_n$. Then, does there exist
some sequence of indices $(i_k),$ for $1 \leq k \leq K$ and for
  $K\geq 1$, and with $1 \leq i_k \leq n$ for all $k$ such that
  \[
    a_{i_1}\dots a_{i_K} = b_{i_1}\dots b_{i_K}?
  \]
  \begin{proof}[Solution]
    This problem is shown to be undecidable through a reduction to the halting
    problem. The proof is technical, and involves encoding the computation
    history of a Turing machine on an input in such a way that the encoding satisfies the
    PCP if and only if the Turing machine accepts the input. It is described in
    detail by Sipser in his book \emph{Introduction to the Theory of Computation}
    \cite{sipser13:_introd_theor_comput}.
  \end{proof}
\end{prob}

@@

\begin{prob}[Disjointness]
  \label{prob:disjointness}
  Let $P,Q$ be context-free grammars. Is $L(P)\cap L(Q) = \varnothing$?
  \begin{proof}[Proof of Undecidability]
    Consider the Post correspondence problem for two lists of words $A,B$. We
    construct two context free grammars from these lists as follows:
    \begin{equation*}
      \begin{split}
        G_A \rightarrow  &\;a_11 \\
        &\!\!\!\!\vdots \\
        G_A \rightarrow  &\;a_nn \\
        G_A \rightarrow  &\;a_1G_A1 \\
        &\!\!\!\!\vdots \\
        G_A \rightarrow  &\;a_nG_An \\
      \end{split}
      \qquad
      \begin{split}
        G_B \rightarrow  &\;a_11 \\
        &\!\!\!\!\vdots \\
        G_B \rightarrow  &\;a_nn \\
        G_B \rightarrow  &\;a_1G_B1 \\
        &\!\!\!\!\vdots \\
        G_B \rightarrow  &\;a_nG_Bn \\
      \end{split}
    \end{equation*}
    Then, we can observe that for some string $s$ to exist in both $L(G_A)$ and $L(G_B)$,
    it must be a solution to the Post correspondence for $A,B$. So, if $L(G_A)\cap
    L(G_B)=\varnothing$, then there are no solutions to the Post correspondence
    for $A,B$. So, it follows that the Post correspondence problem is reducible
    to the problem of the disjointness of context-free grammars, so since the
    Post correspondence problem is undecidable, it follows that the disjointness
    problem is undecidable.
  \end{proof}
\end{prob}

\begin{prob}[Universality]
  \label{prob:cfg:universality}
  Let $G$ be some context-free grammar over an alphabet $\Sigma$. Then, is
  \[
    L(G) = \Sigma^*?
  \]
  \begin{proof}[Proof of Undecidability]
    Consider the Post correspondence problem for two lists of words
    $A,B$. Construct their corresponding grammars $G_A,G_B$ as in Problem~\ref{prob:disjointness}. Then, the languages
    of these grammars are not just \acp{CFL}, but are also \aclp{DCFL},
    discussed in detail in Subsection~\ref{sub:dcfl}, since we
    can construct \aclp{DPDA} to accept them. We have that if
    \[\overline{L(G_A)\cap L(G_B)} = \Sigma^*,\] then
    \[L(G_A)\cap L(G_B)=\varnothing.\] Since these are grammars corresponding to
    lists in a Post correspondence problem, it follows
    that \[\overline{L(G_A)\cap L(G_B)} = \Sigma^*\] implies that there is no
    solution to that Post correspondence problem. Additionally, by DeMorgan's Law, we have
    that
    \[
      \overline{L(G_A)\cap L(G_B)} =
      \overline{L(G_A)}\cup\overline{L(G_B)}.
    \]
    Since \acp{DCFL} are closed under complement and \acp{CFL} are closed under
    union, it follows that $ \overline{L(G_A)\cap L(G_B)}$ is a \ac{CFL}, so the
    question of whether it is empty is decidable. So, it follows that the
    problem of universality reduces to Problem~\ref{prob:pcp}, the Post
    correspondence problem, which is undecidable. So, we have that the
    problem of universality is undecidable.
  \end{proof}
\end{prob}

@@

\begin{prob}[Ambiguity]
  Let $G$ be a context-free grammar. Is $G$ ambiguous?
  \begin{proof}[Proof of Undecidability]
    Consider the Post correspondence problem for two lists of words
    $A,B$. Construct their corresponding grammars $G_A,G_B$ as in Problem~\ref{prob:disjointness}. Now, consider the
    grammar
    \[
      G \rightarrow G_A \vert G_B.
    \]
    It follows that the ambiguity of $G$ implies that a solution to the Post
    correspondence problem for $A,B$ exists, so the Post correspondence problem
    reduces to the ambiguity problem, so the ambiguity problem is undecidable.
  \end{proof}
\end{prob}

\begin{prob}
  Let $G$ be a context free grammars. Is $L(G)$ regular?
  \begin{proof}[Proof of Undecidability]
    Note that $\Sigma^*$ is regular, so if $L(G)$ is not regular then it follows
    that $L(G)\neq \Sigma^*$. Additionally, universality is decidable within
    regular languages, because they are closed under complement and emptiness is
    decidable even within context-free grammars. So, we have that
    Problem~\ref{prob:cfg:universality} reduces to the regularity problem. It
    follows that the regularity problem is undecidable.
  \end{proof}
\end{prob}
\begin{prob}[Equality]
  Let $G_1,G_2$ be context free grammars. Is $L(G_1)=L(G_2)$?
  \begin{proof}[Proof of Undecidability]
    Note that $\Sigma^*$ is regular, so it is also a context-free. So,
    we have that Problem~\ref{prob:cfg:universality} reduces to the equality
    problem. It follows that the equality problem is undecidable.
  \end{proof}
\end{prob}
\begin{prob}[Inclusion]
  Let $G_1,G_2$ be context free grammars. Is $L(G_1)\subseteq L(G_2)$?
  \begin{proof}[Proof of Undecidability]
    As before, note that $\Sigma^*$ is regular, so it is also a
    context-free. Additionally, observe that for any grammar $G$, $L(G)\subseteq
    \Sigma^*$. So,
    we have that Problem~\ref{prob:cfg:universality} reduces to the inclusion
    problem, since $\Sigma^* \subseteq L(G) \implies \Sigma^* = L(G)$. It follows that the inclusion problem is undecidable.
  \end{proof}
\end{prob}

\subsection{Problems decidable for DCFLs}
\label{sub:dcfl}
\begin{defn}
  A \ac{DCFL} is a language accepted by a
  \acl{DPDA}. 
\end{defn}
\acp{DCFL} are another important classification of languages, and are a proper
subset of \acp{CFL}. They are discussed in detail by Ginsburg~\cite{ginsburg65:_deter}.

@@

\begin{prob}
  Let $L$ be a \ac{DCFL}. Is $L = \Sigma^*$?
  \begin{proof}[Proof of Decidability]
    We have that $L$ is a \ac{DCFL}, so by Theorem~\ref{dcfl:complement},
    $\overline{L}$ is also a $\ac{DCFL}$.
    We know that whether $\overline{L}$ is empty is decidable, so it follows
    that the question of whether $L = \Sigma^*$ is also decidable.
  \end{proof}
\end{prob}

@@

One of the earliest questions about the structure of the Turing degrees was
Post's Problem, first posed in 1944~\cite{post44:_recur}. Post's Problem asks
whether there is a computably enumerable degree strictly between $0$ and $0'$. It was solved
independently by Friedberg and Mucnik in the late 1950s using a technique
called the ``priority method''~\cite{Friedberg236, muchnik1956unsolvability}. A solution that does not
involve a priority argument was provided by Ku\v{c}era in 1986~\cite{Kucera:1986:APS:22416.22462}.

% \begin{prob}[Post's Problem]  
%   Does there exist a computably enumerable degree strictly between $0$ and $0'$?
%   % Marchenkov 1976
%   % Independently by Friedberg 57, Mucnik 56/58 using priority method
%   % Priority free solution Kucer 1986''
%   % ``Natural'' solns don't exist
% \end{prob}

@@

\begin{lem}
  \label{lem:deg_count}
  Every Turing degree is of cardinality $\aleph_0$.
  \begin{proof}
    We have that since every Turing machine can be expressed as a finite string over
    $\{0,1\}$, it can be expressed as a finite subset of $\mathbb{N}$. So, it follows
    that for any given oracle, there are at most $\aleph_0$ oracle machines, so
    at most $\aleph_0$ problems are decidable for any Turing degree.
    \par
    Now that an upper bound has been established, we show that the cardinality
    of each degree is exactly $\aleph_0$. Suppose, for the sake of contradiction,
    that a Turing degree contained only a finite number of decision
    problems. Then, the degree would be decidable, since every finite set is
    decidable. But the only decidable degree is $0$, and in the previous section we
    showed that the cardinality of $0$ is $\aleph_0$, so it follows that every
    degree has cardinality $\aleph_0$.
  \end{proof}
\end{lem}

@@

\section{Further Reading}
The study of the Turing degrees was a very active topic in the late
20\textsuperscript{th} and early 21\textsuperscript{st} centuries, and this
paper has just scratched the surface of the field. \par
Readers particularly interested in the structure of the computably enumerable
degrees might be interested in a result by Sacks in 1964 showing their
density~\cite{sacks64:_recur_enumer_degrees_dense}, while those interested in
\aclp{CFG} or the Turing degrees of problems about them should consult Reedy's
1975 paper on the Turing degree of a specific problem~\cite{REEDY197577} and
Ginsburg's book \emph{The Mathematical Theory of Context Free
  Languages}~\cite{ginsburg66:_mathem_theor_contex_languag}. Additionally, a
simple overview of the field is available in Hopcroft's classic text on
computability
theory~\cite[chs.~5--7]{hopcroft07:_introd_autom_theor_languag_comput}. \par For excellent, comprehensive,
and more rigorous textbooks on the study of Turing computability and the Turing
degrees, consider Soare and Lerman's
books~\cite{soare16_turin_comput,lerman16:_degrees_unsol}, as well as
Kleene's original text on the subject~\cite{kleene80_introd}. \par
Finally, Klaus Ambos-Spies and Peter Fejer's unpublished \emph{Degrees
  of Unsolvability} provides an exhaustive overview of the history of the study
of the Turing degrees~\cite{ambos-spies06:_degrees_unsol}.
